---
title: "PPIpriority"
author:
- name: Daniele Pittari
  affiliation: Polytechnic University of Milan, University of Milan
  email: danielepittari@gmail.com
package: PPIpriority
output:
  BiocStyle::pdf_document

vignette: |
  %\VignetteIndexEntry{PPIpriority}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```
```{r style, echo = FALSE, results = 'asis'} 
library(BiocStyle)
```
```{r knitr, echo = FALSE} 
library(knitr)
```

# Introduction

In the last twenty years, the technological advancement of high throughput 
approaches has revolutionized the field of bio medical research. With the 
greater availability of complex data information, the development of tools 
able to model data and extrapolate information has become of great interest.

In this context, graph modelling and network analysis have become popular 
approaches for modelling molecular interactions. Gene co-expression networks
and protein-protein interaction (PPI) networks represent valuable models for 
performing target prioritization of genes/proteins of interest.

In this package, we present a tool which utilizes the STRINGdb PPI 
data sets and allows for target prioritization based on a random walk in the 
PPI from genes of interest (seeds), known for being associated with the 
phenotype/biological process of interest (seeds). 


# Methods

Methodological studies benchmarking the performance of different strategies
for disease gene prioritization in PPI networks have demonstrated that 
approaches utilizing random walking with restart showed good performance in 
the different conditions tested (Köhler et al. 2008). 

Our tool allows to conveniently combine the retrieval of PPI data from the 
STRINGdb (via the homonym package) and conveniently perfom target 
prioritization using a random walking (RW) with restart from the RANKS package.

The output of our package consists of a data.frame containing a ranking of 
the genes based on the RW.


# Workflow
The workflow of PPIpriority is a two-step process

1. Retrieve the PPI network of interest from the STRINGdb database, utilizing
   the function "build_ppi_network" and creating a "PPIpriority_obj" which
   will store your network as well as the STRINGdb object. This allows the user
   to continue data analysis on the downloaded PPI via the STRINGdb package.
   
2. Perform target prioritization on the network stored in the PPIpriority_obj
   via using the function "prioritize_targets" with at least gene as seed
   for the random walk.
    


# Showcase

### Prioritizing putative targets for a protein of interest: the case of ERp44

The first example shows a case of target prioritization on a human PPI.

Given our protein of interest ERp44, we want to prioritize putative functional
targets based on PPI. 

ERP44 is a protein chaperone that cycles from the Endoplasmic
Reticulum (ER) to the cis-Golgi and vice-versa. The role of this protein is to 
selectively retrieve to the ER those proteins that have erroneously left the 
it through vesicle transport. Among the ERP44 binding partners we have 
ER-localized enzymes (e.g. PRDX4, ERAP) or subunits of oligomeric protein 
complexes (e.g. IgMs).


First step: we retrieve the version "11.5" of the STRINGdb PPI for the human 
species (9606) and set a threshold of interaction confidence of 600.

To do this, simply use "build_ppi_network" with 
appropriate inputs for STRINGdb:

```{r Retrieve PPI network}
library(PPIpriority)
#Set  extended timeout, not necessary while using the package
options(timeout=700)

network_ppi_obj<- build_ppi_network(version ="11.5", species = 9606,
score_threshold=700, input_directory=getwd())
```
Now we run target prioritization: to do this, use "prioritize_targets" on your
PPIpriority object. The function will require one or more seed genes that will
be used to initialize the random walk. In our case, ERp44 will be our seed. 
In addition, if you want to analyze a list of candidate genes to prioritize, 
you can insert them as "candidates". The function will pre-filter
the results just based on the genes of interests.

In this scenario, since we are interested in all of the possible results, we 
will not provide candidate genes among "prioritize_targets" inputs. We are 
just passing the HUGO gene name of ERP44, which is ERP44 indeed.


```{r Run prioritization}
output_df<-prioritize_targets(network_ppi_obj, seeds = "ERP44")

head(output_df,15) # show the df
```


ERO1L/ERO1LB, PRDX4, SUMF1, ERAP1 are among the top candidates.
All these proteins have been experimentally demonstrated to physically interact 
with ERp44 (Anelli et al. 2002, Fraldi et al. 2008, Hisatsune et al. 2015, 
Yang et al. 2016). 

As for what concerns the output data.frame: the tool returns the preferred id 
for your input genes according to STRINGdb. 


In these examples we are using HUGO symbols as inputs for seeds (or 
candidates, if any). We suggest to use recommended gene symbol formats.
You can also utilize other formats (e.g. NCBI Entrez Gene), PPIpriority will
convert them into STRINGdb IDs. To achieve better conversion of genes into 
STRINGdb IDs, the mapping step is performed by calling the map method 
of STRINGdb. Check the STRINGdb package for more information.

### Performing the same analysis NCBI Entrez Gene for ERP44 
```{r Run prioritization with ENTREZ gene id}
output_df_erp44_entrez<-prioritize_targets(network_ppi_obj, seeds = "23071")

# is the output equal to out previous dataset obtained with HUGO symbols?
all.equal(output_df_erp44_entrez,output_df)
```

### Prioritizing a list of multiple candidates, given multiple known genes: the case of the functional partners ERp44, LMAN1 and MCFD2
This case concerns the prioritization of candidate targets given multiple 
seeds.

In plasma cells, ERP44, LMAN1 and MCFD2 have been proposed to create 
functional structures in the ER that may facilitate the assembly of complex 
oligomeric proteins. 

Given that we know that coagulation factor V (F5) is a binder of this complex,
we run the analysis with other secreted proteins to prioritize to evalutate if
the algorithm scores F5 among the top ranked gene.


```{r Run prioritization, multiple seeds and candidates}
seeds_genes<-c("ERP44","LMAN1", "MCFD2")
candidate_genes<-c("AREG", "IL1A", "IL1B", "MMP3", "PRL", "F5")

output_df_msmc<-prioritize_targets(network_ppi_obj, 
                              seeds = seeds_genes,
                              candidates=candidate_genes)

# show the df
output_df_msmc
```

The analysis correctly ranks coagulation factor V (F5) at the top in the
ranks. 



# Beyond PPIpriority: continuing your analysis with STRINGdb or other tools for network analysis
The PPIpriority package allows for extrapolating the network modeled during the
"build_ppi_network" step as well as the STRINGdb object. This can be performed 
by using the methods "get_network" and "get_STRINGdb_connection" methods 
respectively. These methods are accessors to the PPIpriority obj S4 
class slots "graph" and "STRINGdb_connection".

### Retrieve the network from PPIpriority obj:

```{r Access netowrk information}
get_network(network_ppi_obj)
```
### Retrieve the STRINGdb obj from PPIpriority obj:
```{r Access STRINGdb obj}
get_STRINGdb_connection(network_ppi_obj)
```


# References
> Köhler S, Bauer S, Horn D, Robinson PN. Walking the interactome for prioritization of candidate disease genes. Am J Hum Genet. 2008 Apr;82(4):949-58. doi: 10.1016/j.ajhg.2008.02.013. Epub 2008 Mar 27. PMID: 18371930; PMCID: PMC2427257.

>Anelli T, Alessio M, Mezghrani A, Simmen T, Talamo F, Bachi A, Sitia R. ERp44, a novel endoplasmic reticulum folding assistant of the thioredoxin family. EMBO J. 2002 Feb 15;21(4):835-44. doi: 10.1093/emboj/21.4.835. PMID: 11847130; PMCID: PMC125352.

>Fraldi A, Zito E, Annunziata F, Lombardi A, Cozzolino M, Monti M, Spampanato C, Ballabio A, Pucci P, Sitia R, Cosma MP. Multistep, sequential control of the trafficking and function of the multiple sulfatase deficiency gene product, SUMF1 by PDI, ERGIC-53 and ERp44. Hum Mol Genet. 2008 Sep 1;17(17):2610-21. doi: 10.1093/hmg/ddn161. Epub 2008 May 28. PMID: 18508857.

>Hisatsune C, Ebisui E, Usui M, Ogawa N, Suzuki A, Mataga N, Takahashi-Iwanaga H, Mikoshiba K. ERp44 Exerts Redox-Dependent Control of Blood Pressure at the ER. Mol Cell. 2015 Jun 18;58(6):1015-27. doi: 10.1016/j.molcel.2015.04.008. Epub 2015 May 7. PMID: 25959394.

>Yang K, Li DF, Wang X, Liang J, Sitia R, Wang CC, Wang X. Crystal Structure of the ERp44-Peroxiredoxin 4 Complex Reveals the Molecular Mechanisms of Thiol-Mediated Protein Retention. Structure. 2016 Oct 4;24(10):1755-1765. doi: 10.1016/j.str.2016.08.002. Epub 2016 Sep 15. PMID: 27642162.
  
# Session information
```{r sessionInfo, echo=FALSE}
sessionInfo()
```
